First Order Alternation

### 2019-02-05 quantifiers/body.tex
ok

\begin{definition}\label{def:unfolding}
  An \emph{unfolding} of an automaton $\A =
  \tuple{\Sigma,X,Q,\iota,F,\Delta}$ is a finite partial mapping $U :
  \Sigma^* \rightharpoonup_{\mathit{fin}} \posforms(Q,\emptyset)$,
  whose domain $\dom(U)$ is a finite prefix-closed complete set, such
  that $U(\epsilon) = \iota$, and for each sequence $\alpha a \in
  \dom(U)$, such that $\alpha \in \Sigma^*$ and $a \in \Sigma$:
  \[\begin{array}{c}
  \stamp{U(\alpha)}{0} \wedge
  \bigwedge_{q(\vec{y}) \arrow{a(X)}{} \psi} \forall y_1 \ldots
  \forall y_{\#q} ~.~ \stamp{q}{0}(\vec{y}) \rightarrow
  \stamp{\psi}{1} \models \stamp{U(\alpha a)}{1}
  \end{array}\]  
  $U$ is \emph{safe} if for each $\alpha \in \dom(U)$, the formula
  $U(\alpha) \wedge \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots
  \forall_{y_{\#(q)}} ~.~ q(\vec{y}) \rightarrow \bot$ is
  unsatisfiable.
\end{definition}

@@

\begin{proposition}\label{prop:local-gli}
  If there exists a Lyndon interpolant for any two formulae $\phi$ and
  $\psi$, in the first order theory of data with uninterpreted
  predicate symbols, such that $\phi \wedge \psi$ is unsatisfiable,
  then any sequence of input events $\alpha = a_1 \ldots a_n \in
  \Sigma^*$, such that $\accform{\alpha}$ is unsatisfiable, has a
  local GLI $(I_0,\ldots,I_n)$.
\end{proposition}
\paragraph{\em Proof.} By definition, $\accform{\alpha}$ is the formula:
\[\begin{array}{c}
\stamp{\iota}{0} \wedge 
 \bigwedge_{i=1}^n \bigwedge_{q(\vec{y}) \arrow{a_i(X)}{} \psi \in
   \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
 \stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i} \wedge
 \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)}
 ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot
 \end{array}\]
 We define the formulae: 
 \[\begin{array}{rcl}
 \varphi_i & \isdef & \bigwedge_{q(\vec{y}) \arrow{a_i(X)}{} \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
 \stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i},~ \text{for all}~ i \in [1,n] \\
 \psi & \isdef & \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)}
 ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot
 \end{array}\]
 Observe that $\voc{\stamp{\iota}{0}} \subseteq \stamp{Q}{0}$,
 $\voc{\varphi_i} \subseteq
 \stamp{Q}{i-1}\cup\stamp{Q}{i}\cup\stamp{X}{i}$, for all $i \in
       [1,n]$, and $\voc{\psi} \subseteq \stamp{Q}{n}$.  We apply
       Lyndon's Interpolation Theorem for the formulae
       $\stamp{\iota}{0}$ and $\bigwedge_{i=1}^n \varphi_i \wedge
       \psi$ and obtain a formula $I_0$, such that $\stamp{\iota}{0}
       \models I_0$, $I_0 \wedge \bigwedge_{i=1}^n \varphi_i \wedge
       \psi$ is unsatisfiable, $\voc{I_0} \subseteq
       \voc{\stamp{\iota}{0}} \cap (\bigcup_{i=1}^n \voc{\varphi_i}
       \cup \voc{\psi}) \subseteq \stamp{Q}{0}$ and $\pset{-}{I_0}
       \subseteq \pset{-}{\stamp{\iota}{0}} \cap (\bigcup_{i=1}^n
       \pset{-}{\varphi} \cup \pset{-}{\psi}) = \emptyset$. Repeating
       the reasoning for the formulae $I_0 \wedge \varphi_1$ and
       $\bigwedge_{i=2}^n \varphi_i \wedge \psi$, we obtain $I_1$,
       such that $I_0 \wedge \varphi_1 \models I_1$, $I_1 \wedge
       \bigwedge_{i=2}^n \varphi_i \wedge \psi$ is unsatisfiable,
       $\voc{I_1} \subseteq (\voc{I_0} \cup \voc{\varphi_1}) \cap
       (\bigcup_{i=2}^n \voc{\varphi_i} \cup \voc{\psi}) \subseteq
       \stamp{Q}{1}$ and $\pset{-}{I_1} \subseteq (\pset{-}{I_0} \cup
       \pset{-}{\varphi_1}) \cap (\bigcup_{i=2}^n \pset{-}{\varphi_i}
       \cup \pset{-}{\psi}) = \emptyset$. Continuing in this way, we
       obtain formulae $I_0, I_1, \ldots, I_{n}$ as required. \qed

@@

\begin{lemma}\label{lemma:refinement}
  Let $U$ be an unfolding of an automaton $\A =
  \tuple{\Sigma,X,Q,\iota,F,\Delta}$ such that $\alpha = a_1\ldots a_n
  \in \dom(U)$ is an input event sequence and $(I_0, \ldots, I_n)$ be
  a GLI for $\alpha$. Then the following mapping $U' : \dom(U)
  \rightarrow \posforms(Q,\emptyset)$ is an unfolding of $\A$, where:
  \begin{compactitem}
  \item $U'(\alpha_k) = U(\alpha_k) \wedge J_k$, for all $k \in [n]$,
    where $J_k$ is the formula obtained from $I_k$ by replacing each
    time-stamped predicate symbol $\stamp{q}{k}$ by $q$ and
    existentially quantifying each free variable in $I_k$,
  %
\item $U'(\beta) = U(\beta)$ if $\beta \in \dom(U)$ and $\beta
  \not\prefix \alpha$,
  \end{compactitem}
\end{lemma}
\proof{
  The new set of formulae $U'(\alpha_0), \ldots, U'(\alpha_n)$ complies
  with Definition \ref{def:unfolding}, because: \begin{compactitem}
  \item $U'(\alpha_0) \equiv \iota$, since, by point
    \ref{it2:generalized-lyndon-interpolant} of Definition
    \ref{def:generalized-lyndon-interpolant}, we have $\stamp{\iota}{0}
    \models I_0$, thus $\iota \models J_0$ and $U'(\alpha_0) =
    U(\alpha_0) \wedge J_0 \equiv \iota \wedge J_0 \equiv \iota$, and
    %
  \item by Definition \ref{def:generalized-lyndon-interpolant}
    (\ref{it3:generalized-lyndon-interpolant}) we have, for all $k \in
    [n-1]$:
    \[I_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
    \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{k}(\vec{y})
    \rightarrow \stamp{\psi}{k+1} \models I_{k+1}\] We write
    $\overstamp{I}{j}_k$ for the formula in which each predicate
    symbol $\stamp{q}{k}$ is replaced by $\stamp{q}{j}$. Then the
    following entailment holds:
    \[\overstamp{I}{0}_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
    \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
    \rightarrow \stamp{\psi}{1} \models \overstamp{I}{1}_{k+1}\]
    Because $J_k$ is obtained by removing the time stamps from the predicate symbols and  
    existentially quantifying all the free variables of $I_k$, we also
    obtain, applying Fact \ref{fact:ex-entail} below:
    \[\stamp{J}{0}_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
    \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
    \rightarrow \stamp{\psi}{1} \models \stamp{J}{1}_{k+1}\] Since $U$
    satisfies the labeling condition of Definition \ref{def:unfolding}
    (\ref{it3:unfolding}) and $U'(\alpha_k) = U(\alpha_k) \wedge J_k$,
    we obtain, as required: 
    \[\stamp{U'(\alpha_k)}{0} \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
    \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
    \rightarrow \stamp{\psi}{1} \models
    \stamp{U'(\alpha_{k+1})}{1}\enspace.\]
  \end{compactitem}

@@

Observe that, by Lemma \ref{lemma:subst-quant}
(\ref{it2:subst-quant}), the set of free variables of a GLI formula
$I_k$ consists of \begin{inparaenum}[(i)]
\item variables $\stamp{X}{\leq k}$ keeping track of data values seen
  in the input at some earlier moment in time, and
%
\item variables that track past choices made within the transition
  rules.
\end{inparaenum}
Basically, it is not important when exactly in the past a certain
input has been read or when a choice has been made, as only the
relation between the values of these and the current variables
determines the future behavior of the automaton. Intuitively,
quantifying these variables existentially does the job of ignoring
when exactly in the past these values have been seen.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\enlargethispage{6mm}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\vspace*{-\baselineskip}
\section{Conclusions}

>>>>

\begin{definition}\label{def:unfolding}
  An \emph{unfolding} of an automaton $\A =
  \tuple{\Sigma,X,Q,\iota,F,\Delta}$ is a finite partial mapping $U :
  \Sigma^* \rightharpoonup_{\mathit{fin}} \posforms(Q,\emptyset)$,
  whose domain $\dom(U)$ is a finite prefix-closed complete set, such
  that $U(\epsilon) = \iota$, and for each sequence $\alpha a \in
  \dom(U)$, such that $\alpha \in \Sigma^*$ and $a \in \Sigma$:
  \[\begin{array}{c}
  \stamp{U(\alpha)}{0} \wedge
  \bigwedge_{q(\vec{y}) \arrow{a(X)}{} \psi} \forall y_1 \ldots
  \forall y_{\#q} ~.~ \stamp{q}{0}(\vec{y}) \rightarrow
  \stamp{\psi}{1} \models \stamp{U(\alpha a)}{1}
  \end{array}\]
  A path $\alpha$ is \emph{safe} in $U$ if and only if $U(\alpha)
  \wedge \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots
  \forall_{y_{\#(q)}} ~.~ q(\vec{y}) \rightarrow \bot$ is
  unsatisfiable. The unfolding $U$ is safe if and only if every path
  in $\dom(U)$ is safe in $U$.
\end{definition}

@@

\begin{proposition}\label{prop:local-gli}
  If there exists a Lyndon interpolant for any two formulae $\phi$ and
  $\psi$, in the first order theory of data with uninterpreted
  predicate symbols, such that $\phi \wedge \psi$ is unsatisfiable,
  then any sequence of input events $\alpha = a_1 \ldots a_n \in
  \Sigma^*$, such that $\accform{\alpha}$ is unsatisfiable, has a
  local GLI $(I_0,\ldots,I_n)$.
\end{proposition}
\paragraph{\em Proof.} By definition, $\accform{\alpha}$ is the formula:
\[\begin{array}{c}
\stamp{\iota}{0} \wedge 
 \bigwedge_{i=1}^n \underbrace{\bigwedge_{q(\vec{y}) \arrow{a_i(X)}{} \psi \in
   \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
 \stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i}}_{\varphi_i} \wedge
 \underbrace{\bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)}
 ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot}_{\phi}
 \end{array}\]
 %% We define the formulae: 
 %% \[\begin{array}{rcl}
 %% \varphi_i & \isdef & \bigwedge_{q(\vec{y}) \arrow{a_i(X)}{} \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
 %% \stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i},~ \text{for all}~ i \in [1,n] \\
 %% \phi & \isdef & \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)}
 %% ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot
 %% \end{array}\]
 Observe that $\voc{\stamp{\iota}{0}} \subseteq \stamp{Q}{0}$,
 $\voc{\varphi_i} \subseteq
 \stamp{Q}{i-1}\cup\stamp{Q}{i}\cup\stamp{X}{i}$, for all $i \in
       [1,n]$, and $\voc{\phi} \subseteq \stamp{Q}{n}$.  We apply
       Lyndon's Interpolation Theorem for the formulae
       $\stamp{\iota}{0}$ and $\bigwedge_{i=1}^n \varphi_i \wedge
       \phi$ and obtain a formula $I_0$, such that $\stamp{\iota}{0}
       \models I_0$, $I_0 \wedge \bigwedge_{i=1}^n \varphi_i \wedge
       \phi$ is unsatisfiable, $\voc{I_0} \subseteq
       \voc{\stamp{\iota}{0}} \cap (\bigcup_{i=1}^n \voc{\varphi_i}
       \cup \voc{\phi}) \subseteq \stamp{Q}{0}$ and $\pset{-}{I_0}
       \subseteq \pset{-}{\stamp{\iota}{0}} \cap (\bigcup_{i=1}^n
       \pset{-}{\varphi} \cup \pset{-}{\phi}) = \emptyset$. Repeating
       the reasoning for the formulae $I_0 \wedge \varphi_1$ and
       $\bigwedge_{i=2}^n \varphi_i \wedge \phi$, we obtain $I_1$,
       such that $I_0 \wedge \varphi_1 \models I_1$, $I_1 \wedge
       \bigwedge_{i=2}^n \varphi_i \wedge \phi$ is unsatisfiable,
       $\voc{I_1} \subseteq (\voc{I_0} \cup \voc{\varphi_1}) \cap
       (\bigcup_{i=2}^n \voc{\varphi_i} \cup \voc{\phi}) \subseteq
       \stamp{Q}{1}$ and $\pset{-}{I_1} \subseteq (\pset{-}{I_0} \cup
       \pset{-}{\varphi_1}) \cap (\bigcup_{i=2}^n \pset{-}{\varphi_i}
       \cup \pset{-}{\phi}) = \emptyset$. Continuing in this way, we
       obtain formulae $I_0, I_1, \ldots, I_{n}$ as required. \qed

@@

\begin{lemma}\label{lemma:refinement}
  Let $U$ be an unfolding of an automaton $\A =
  \tuple{\Sigma,X,Q,\iota,F,\Delta}$ such that $\alpha = a_1\ldots a_n
  \in \dom(U)$ is an input event sequence and $(I_0, \ldots, I_n)$ be
  a GLI for $\alpha$. Then the following mapping $U' : \dom(U)
  \rightarrow \posforms(Q,\emptyset)$ is an unfolding of $\A$, where:
  \begin{compactitem}
  \item $U'(\alpha_k) = U(\alpha_k) \wedge J_k$, for all $k \in [n]$,
    where $J_k$ is the formula obtained from $I_k$ by replacing each
    time-stamped predicate symbol $\stamp{q}{k}$ by $q$ and
    existentially quantifying each free variable in $I_k$,
  %
\item $U'(\beta) = U(\beta)$ if $\beta \in \dom(U)$ and $\beta
  \not\prefix \alpha$,
  \end{compactitem}
  Moreover, $\alpha$ is safe in $U'$.
\end{lemma}
\proof{
  The new set of formulae $U'(\alpha_0), \ldots, U'(\alpha_n)$ complies
  with Definition \ref{def:unfolding}, because: \begin{compactitem}
  \item $U'(\alpha_0) \equiv \iota$, since, by point
    \ref{it2:generalized-lyndon-interpolant} of Definition
    \ref{def:generalized-lyndon-interpolant}, we have $\stamp{\iota}{0}
    \models I_0$, thus $\iota \models J_0$ and $U'(\alpha_0) =
    U(\alpha_0) \wedge J_0 \equiv \iota \wedge J_0 \equiv \iota$, and
    %
  \item by Definition \ref{def:generalized-lyndon-interpolant}
    (\ref{it3:generalized-lyndon-interpolant}) we have, for all $k \in
    [n-1]$:
    \[I_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
    \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{k}(\vec{y})
    \rightarrow \stamp{\psi}{k+1} \models I_{k+1}\] We write
    $\overstamp{I}{j}_k$ for the formula in which each predicate
    symbol $\stamp{q}{k}$ is replaced by $\stamp{q}{j}$. Then the
    following entailment holds:
    \[\overstamp{I}{0}_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
    \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
    \rightarrow \stamp{\psi}{1} \models \overstamp{I}{1}_{k+1}\]
    Because $J_k$ is obtained by removing the time stamps from the predicate symbols and  
    existentially quantifying all the free variables of $I_k$, we also
    obtain, applying Fact \ref{fact:ex-entail} below:
    \[\stamp{J}{0}_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
    \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
    \rightarrow \stamp{\psi}{1} \models \stamp{J}{1}_{k+1}\] Since $U$
    satisfies the labeling condition of Definition \ref{def:unfolding}
    (\ref{it3:unfolding}) and $U'(\alpha_k) = U(\alpha_k) \wedge J_k$,
    we obtain, as required: 
    \[\stamp{U'(\alpha_k)}{0} \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
    \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
    \rightarrow \stamp{\psi}{1} \models
    \stamp{U'(\alpha_{k+1})}{1}\enspace.\]
  \end{compactitem}

@@

Observe that, by Lemma \ref{lemma:subst-quant}
(\ref{it2:subst-quant}), the set of free variables of a GLI formula
$I_k$ consists of \begin{inparaenum}[(i)]
\item variables $\stamp{X}{\leq k}$ keeping track of data values seen
  in the input at some earlier moment in time, and
%
\item variables that track past choices made within the transition
  rules.
\end{inparaenum}
Basically, it is not important when exactly in the past a certain
input has been read or when a choice has been made, as only the
relation between the values of these and the current variables
determines the future behavior of the automaton. Intuitively,
quantifying these variables existentially does the job of ignoring
when exactly in the past these values have been seen. Moreover, the
last point of Lemma \ref{lemma:refinement} ensures that the refined
path is safe in the new unfolding and thus will stay safe in all
future unfoldings of $\A$.

@@

@@

\section{Conclusions}
